---
title: "Day 1: Basic data manipulation and visualization"
author: "Nayef Ahmad - VCH Decision Support"
date: "September 14, 2018"
output: 
    html_document: 
        toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library("kableExtra")


# shortcuts: 
# > ctrl + alt + i ==> new code chunk
# > ctrl + shift + k ==> knit doc 
# \  ==> line break (include 2 spaces after)
```


## Why are we talking about R/Python? 
* It's the industry standard...
* ... for good reason 

\  

Ok, let's get started 

*********


## Setup
1. Create a folder where you'll save project files. I recommend following a folder structure similar to the one mentioned [here](https://swcarpentry.github.io/r-novice-gapminder/02-project-intro/) 
2. Open RStudio, create a new project, save it in the root folder. 
3. Open a new R script. Now let's get started with the code! 

\  
\  
\  

## Install the packages you need
You will only need to do this once. I recommend running these lines in the console instead of the script editor. 

```{r eval = FALSE}
install.packages("dplyr")
install.packages("ggplot2")
install.packages("here")


```

## Load the packages for the current session
```{r message = FALSE}
library("dplyr")  # for manipulating data frames 
library("ggplot2")  # for data viz
library("here")  # for simplifying folder references 


```

*******************

\  
\  
\  

## Built-in datasets
R comes with some built-in datasets. Let's start by taking a look at them. 

### mtcars dataset 
```{r}
mtcars %>% kable()

```

### Summary functions: 
These functions are good for getting a quick sense of a dataset. 

```{r}
str(mtcars)
```

```{r}
summary(mtcars)
```

```{r}
head(mtcars)  # by default shows top 10 rows 
```

\  

### Mini-exercise
Try doing the same with the following datasets: iris, airquality, diamonds

\  
\  
\  



