---
title: 'Day 2: Time series analysis and forecasting'
author: "Nayef Ahmad - VCH Decision Support"
date: "November 3, 2018"
output: 
    html_document: 
      toc: yes
      code_folding: hide 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE)


library("kableExtra")


# shortcuts: 
# > ctrl + alt + i ==> new code chunk
# > ctrl + shift + k ==> knit doc 
# \  ==> line break (include 2 spaces after)
```

## Load the packages and functions for the current session
```{r message = FALSE, warning = FALSE}
library(dplyr)  # for manipulating data frames 
library(ggplot2)  # for data viz
library(here)  # for simplifying folder references 
library(readr)  # reading and writing data files 
library(fpp)  # forecasting principles and practices 
library(prophet)  # facebook's forecasting package 
library(janitor)  # for cleaning up dataframes  
library(lubridate)  # for working with dates 
library(tidyr)  # for tidying data
library(magrittr)  # for easy piping 

source(here::here("src", "stl.as.df_function.R"))

```

## Time series analysis 
### Daily parking tickets in Vancouver
#### Read in and clean data for 2017 
```{r}
# if you want the entire parking data set, run this: 
# warning: this will take a long time. 

# df1.1.tickets.2017 <- read.csv(url("ftp://webftp.vancouver.ca/opendata/csv/2017Parking_Tickets.csv"))

# df1.2.tickets.2016 <- read.csv(url("ftp://webftp.vancouver.ca/opendata/csv/2016Parking_Tickets.csv"))



# in case it's easier to load this off disk instead of 
# over network: 
df1.1.tickets.2017 <- read_csv(here::here("data", 
                        "vancouver-parking-tickets-2017.csv")) %>%
    select(-"X1")

df1.2.tickets.2016 <- read_csv(here::here("data", 
                        "vancouver-parking-tickets-2016.csv"))

# union of the two datasets: 
df1.tickets <- rbind(df1.2.tickets.2016, 
                     df1.1.tickets.2017)


# let's do some wrangling: 
df1.tickets %<>% 
    clean_names() %>% 
    mutate(entry_date = mdy(entry_date)) %>% 
    mutate_if(is.character, 
              factor)

# view result: 
# str(df1.tickets)
# summary(df1.tickets)
head(df1.tickets) %>% 
    kable() %>% 
    kable_styling(bootstrap_options = c("striped",
                                        "condensed", 
                                        "responsive"), 
                  full_width = FALSE, 
                  position = "left")


```

#### Group by date
Note that we have to join on list of all dates in order to make sure there are no missing dates in the data.  

```{r}
# group by 
df2.daily.tickets <- df1.tickets %>% 
    group_by(entry_date) %>% 
    summarise(count = n())

# all dates: 
df3.1.all.dates <- data.frame(dates = seq(as.Date("2016-01-01"),
                                        as.Date("2017-12-31"),
                                        by="1 day"))

# full join: 
df3.daily.tickets.joined <- df2.daily.tickets %>% 
    full_join(df3.1.all.dates, 
              by = c("entry_date" = "dates")) %>% 
    arrange(entry_date) %>% 
    
    # replace NAs with 0s: 
    mutate(count = replace_na(count, 0))


# view result: 
# str(df3.daily.tickets.joined)
# summary(df3.daily.tickets.joined)
head(df3.daily.tickets.joined) %>% 
    kable() %>% 
    kable_styling(bootstrap_options = c("striped",
                                        "condensed", 
                                        "responsive"), 
                  full_width = FALSE, 
                  position = "left")

    
tail(df3.daily.tickets.joined) %>% 
    kable() %>% 
    kable_styling(bootstrap_options = c("striped",
                                        "condensed", 
                                        "responsive"), 
                  full_width = FALSE, 
                  position = "left")



```

Interesting that there's no data for Christmas and New Year's.  

\  
\  
\  

#### Visualize time series: 
```{r}
p1.parking.2017 <- df2.daily.tickets %>% 
    ggplot(aes(x = entry_date, 
               y = count)) + 
    geom_line() + 
    labs(title = "Daily parking tickets in Vancouver, BC", 
         subtitle = "2016-01-01 to 2017-12-31\n\n", 
         caption = "\n\nSource:https://data.vancouver.ca/datacatalogue/parking-tickets.htm") + 
    
    geom_smooth() +
    
    theme_classic(base_size = 16) + 
    theme(plot.caption = element_text(size = 11))

p1.parking.2017

```


#### Break down into time series components

First we have to convert the counts into a time series object
```{r}
# reference: https://robjhyndman.com/hyndsight/dailydata/

t1.tickets.time.series <- df3.daily.tickets.joined %>% 
    pull(count) %>%   # pull out the count column as a vector
    ts(frequency = 365, 
       start = c(2016)) 

# t1.tickets.time.series
```


Now we can run an  STL decomposition. 

```{r}
stl(t1.tickets.time.series,
    s.window = "periodic") %>% plot

# get decomposition as a df: 
df4.stl.decomp <- stl.as.df_fn(t1.tickets.time.series)

# join back on orig data: 
df4.stl.decomp <- cbind(df3.daily.tickets.joined, 
                        df4.stl.decomp) %>% 
    select(-"data")

# str(df4.stl.decomp)
# summary(df4.stl.decomp)
head(df4.stl.decomp, 20) %>% 
    kable() %>% 
    kable_styling(bootstrap_options = c("striped",
                                        "condensed", 
                                        "responsive"), 
                  full_width = FALSE, 
                  position = "left")

```

It's probably easier to explain these components with a simpler case, so we'll look at the the standard textbook example:  


```{r}
plot(elecequip)

stl(elecequip, 
    s.window = "periodic") %>% plot

stl.as.df_fn(elecequip) %>% head

```














## Producing forecasts 










## Evaluating forecasts 
