---
title: 'Day 4: Generalized linear models'
author: "Nayef Ahmad - VCH Decision Support"
date: "September 26, 2018"
output: 
    html_document: 
      toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE)


library("kableExtra")


# shortcuts: 
# > ctrl + alt + i ==> new code chunk
# > ctrl + shift + k ==> knit doc 
# \  ==> line break (include 2 spaces after)
```


## Extending the linear model using link functions 
The linear models that we have looked at so far have 2 basic components. Recall the Boston house prices model we worked on last session. The components of the model we built were: 

1. Random component: the y-variable 
2. Linear predicto: the x-variables and the coefficients associated with them (betas) 

We're now going to move to a more general type of model, which has three components: 
1. Random component 
2. Linear predictor
3. Link function 

Todo: image of link function, etc. 

***********************************

\  
\  
\  

## Load the packages for the current session
```{r message = FALSE}
library("dplyr")  # for manipulating data frames 
library("ggplot2")  # for data viz
library("here")  # for simplifying folder references 
library("readr")  # reading and writing data files 
library("GGally")  # extension to ggplot2
library("ISLR")  # package from the book Intro to Stat Learning 

```

*******************

## Examining the Default dataset
The default dataset is built into the ISLR package. 

```{r}
df1.default.dataset <- Default  # rename for convenience 

str(df1.default.dataset)

levels(df1.default.dataset$student)  # No is the first level, 1
levels(df1.default.dataset$default)  # No is the first level, 1

```

#### Quick summaries 
```{r}
summary(df1.default.dataset)
```

```{r}
head(df1.default.dataset) %>% 
    # don't use the following lines; they're specifically for Rmarkdown files
    kable() %>% 
    kable_styling(bootstrap_options = c("striped",
                                        "condensed", 
                                        "responsive"), 
                  full_width = FALSE, 
                  position = "left")
```


#### Pairs plot 

```{r}
p1.pairs <- df1.default.dataset %>% 
    ggpairs(); p1.pairs


    
```

There's a lot of interesting things to see here. 

1. Defaults are a small proportion of total accounts, for both students and non-students. Todo: how different are the proportions? 
2. Non-students have much higher income than students (no surprise). They also have slightly higher credit balances. 
3. todo: ...





```{r}

df1.default.dataset %>% 
    group_by(student) %>% 
    summarize(avg.bal = mean(balance), 
              avg.inc = mean(income), 
              num = n())

```


## A terrible classifier: just use averages 

```{r}

df1.default.dataset %>% 
    group_by(default) %>% 
    summarize(avg.bal = mean(balance), 
              avg.inc = mean(income), 
              num.cases = n(), 
              num.student = sum(student=="Yes"), 
              perc.student = round(100*sum(student=="Yes")/n(), 2)) %>% 
    
    
    kable() %>% 
    kable_styling(bootstrap_options = c("striped",
                                        "condensed", 
                                        "responsive"), 
                  full_width = FALSE, 
                  position = "left")

```

